---
title: "Immigrant Apprehensions in the US' Southwest Border"
subtitle: "A Time Series Analysis of Migration Trends (2000–2025)"
author: "Paola Pavón"
date: "2025-04-08"
output: 
  html_document:
    toc: true

---
[View this report online](https://mellow-lily-332675.netlify.app/)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dygraphs) # For cute charts
library(zoo) #For moving averages
library(forecast) # To compare between years
library(knitr) # For tables
library(htmltools) # To knit properly
```

## Introduction

Migration at the U.S. Southern Border is a complex and highly debated issue, 
shaped by a mix of political, economic, social, and environmental factors. 
In recent decades, the number of migrant encounters at the Southwest Border 
has fluctuated in response to both structural pressures in sending countries as
well as changes in U.S. immigration policy. Understanding these fluctuations 
calls for a rigorous, data-driven analysis grounded in time series methods.

This report investigates the evolution of migrant apprehensions at the 
U.S.–Mexico border from 2000 to 2025, with a particular focus on identifying
long-term trends, seasonal patterns, and structural changes across different
presidential administrations. The analysis also explores how apprehensions
relate to conditions in Mexico, including economic activity and insecurity, and
climate in El Paso, Texas. By applying classical time series tools such as 
decomposition and correlation analysis, the goal is to discover the dynamics
underlying migration patterns and assess how they respond to both domestic and
international influences.

## Obtaining the data

The data collection process revealed important considerations and limitations 
that shaped the structure of this analysis. Border apprehension data has been 
recorded annually since the 1960s, but it wasn’t until October 1999 that the 
U.S. Customs and Border Protection (CBP) began reporting at a monthly 
resolution. Therefore, this marks the beginning of our time series.

Additionally, the fiscal year used by the U.S. Customs Service does not align
with the calendar year. For example, the “2000” report actually reflects data 
from October 1999 to September 2000. This required special care during data 
cleaning and alignment, as something reported like "October 2000" actually meant
October 1999.

After September 2019, a shift in reporting style introduced new categories such
as *inadmissibles, expulsions*, and *apprehensions*. This project focuses 
exclusively on the last two.

Finally, the dataset ends in February 2025, which means the most recent policy 
impacts (e.g., from the 2024 U.S. election) are not yet captured.

### Why Focus on Mexico?
Mexico accounts for the largest share of migrant apprehensions at the U.S. 
Southwest Border in recent years. According to official encounter statistics 
between 2019 and 2022, over 2 million Mexican nationals were apprehended — 
significantly more than any other nationality. The latter justifies a focused 
analysis on Mexican variables such as economic activity and public security to
understand how these internal conditions may shape migratory pressure. While 
migration from Central America and other regions has grown, Mexico remains the 
most consistently represented group in border enforcement data.

```{r, echo = FALSE}
apprehensions <- data.frame(
  Country = c("Mexico", "Guatemala", "Honduras", "El Salvador", "Other", 
              "Unknown"),
  `2019` = c(166458, 185535, 190872, 71501, 216569, 21073),
  `2020` = c(131057, 41512, 39327, 13269, 72991, 452),
  `2021` = c(608037, 283771, 309174, 93780, 262249, 1332),
  `2022` = c(1522740, 228258, 221684, 53787, 379282, 585),
  stringsAsFactors = FALSE,
  check.names = FALSE
)

# Add Grand Total column
apprehensions$GrandTotal <- rowSums(apprehensions[ , 2:5])

apprehensions_formatted <- apprehensions
apprehensions_formatted[ , 2:6] <- lapply(apprehensions_formatted[ , 2:6], 
                    function(x) format(x, big.mark = ",", scientific = FALSE))


kable(apprehensions_formatted, caption = "Table 1: U.S. Border Apprehensions by Country 
      of Origin (2019–2022)", align = "lrrrrr", format = "markdown")
```

## Creating the time series

Once the data was cleaned and aligned across years, we proceeded to create the 
time series using R’s ts() function. This allowed us to explore the evolution of
monthly apprehensions at the U.S. Southwest Border over more than two decades.

Our first time series plot already communicated several key insights:

**Figure 1.** *Monthly Migrant Apprehensions at the U.S. Southwest Border
(2000–2025)*
```{r, echo=FALSE}
# Importing data
setwd(getwd())
file_path <- file.path("Data", "Clean Data.csv")
data <- read.csv(file_path, header = TRUE)
colnames(data) <- c("Date", "Encounters")

# Date formating 
data$Date <- as.Date(data$Date, format="%d/%m/%Y") 
data$Date <- as.Date(data$Date,format="%Y/%m/%d" )
data <- data[-(1:3),]

# TS object creation
FirstYear <- format(data$Date[1], "%Y")
FirstMonth <- format(data$Date[1], "%m")
apprehensionsTS <- ts(data$Encounters, start=c(FirstYear, FirstMonth), 
                    frequency=12)
# Let's plot it!
tags$div(
  style = "text-align: center;",
  dygraph(apprehensionsTS) |> dyRangeSelector()
)

```

- **A clear long-term trend:** Apprehensions steadily declined from the early 
2000s until around 2011. However, from 2017 onward, the trend reversed 
dramatically. Our first guess was that this change followed Trump's 
anti-migration policies.

- **Seasonality:** A regular, recurring spike can be observed almost every year 
— most notably around the spring months, which we later confirm aligns with
March.

- **Sharp increases in recent years:** Particularly after 2019, apprehension 
levels surged, peaking around 2022–2023.

- **Historical lows and highs:** The lowest point occurs in April 2020,
likely due to the start of the COVID-19 pandemic. Likewise, late 2023 marks 
the highest number of apprehensions in the entire dataset.

## Trend Analysis

The chart below visualizes the monthly total migrant apprehensions at the U.S.
Southwest Border from 2000 to early 2025. The black line shows raw monthly data,
while the red line represents a 12-month moving average, helping us identify 
underlying trends.

**Figure 2.** *Monthly Migrant Apprehensions with 12-Month Moving Average* 
```{r Trend, fig.align = 'center', fig.width=8, fig.height=5, echo = FALSE }
plot(apprehensionsTS, ylab= "Number of Encounters")
averages = rollmean(apprehensionsTS, k=12, align="right")
lines(averages, col="red") 
```

Several key patterns emerge from this graph:

- **Early 2000s Decline:** Apprehensions steadily decreased from 2000 to around 
2011. This could reflect improved border enforcement, declining Mexican 
migration due to demographic and economic changes, or shifts in U.S. policy.

- **Stability (2012–2017):** Apprehensions remained relatively low and stable
during this period. This includes much of the Obama administration's second term,
where enforcement continued but without dramatic new measures.

- **Sharp Increase Post-2017:** From 2017 onward, apprehensions grew
significantly. This aligns with the beginning of Donald Trump’s presidency and 
his administration’s restrictive immigration policies. Rather than discouraging
migration, some policies (e.g. *Remain in Mexico*) may have led to increased 
border crossings under worsening conditions.

- **Pandemic Dip in 2020:** The steep decline in early 2020 corresponds with the 
onset of COVID-19 and temporary border closures. The immediate drop highlights 
how public health emergencies can disrupt migration flows.

- **Post-2020 Spike:** From mid-2020 through 2023, apprehensions reached 
historic highs — particularly under President Biden’s term. This could reflect a
combination of higher migration demand after pandemic disruptions, shifting
enforcement practices, and worsening conditions in migrants' home countries.

This long-term view shows that while migration patterns can be influenced by
policy, they are also shaped by broader social, economic, and environmental
forces — many of which transcend national boundaries.

## Seasonality Analysis
To better understand the underlying structure of migrant apprehension data, we 
applied the decompose() function in R. This function separates a time series into
three key components:

- **Trend:** The long-term movement in the data, ignoring short-term
fluctuations.

- **Seasonal:** The recurring, systematic patterns that repeat over a fixed 
period (e.g., every 12 months).

- **Random (or Residual):** The irregular, unexplained variation remaining after
removing the trend and seasonal patterns.

**Figure 3.** *Time Series Decomposition of Apprehensions*
```{r seasonality, fig.align = 'center',fig.width=8, fig.height=5, echo = FALSE}
decomposition <- decompose(apprehensionsTS)
plot(decomposition) 
```
The plot generated by this function provided crucial insights:

- The **trend** component reaffirmed what we observed visually — a downward
trend until around 2011, followed by a surge beginning in 2017.

- The **seasonal** component revealed consistent spikes in the spring months, 
especially March. This confirmed our earlier hypothesis regarding weather-driven
migration timing.

- The **random** component captured events like the dip in 2020, which are not
part of seasonal or trend behavior.

The strength of this approach is that it helps isolate what’s systematic vs.
what’s unexpected, allowing for clearer interpretation of the driving forces 
behind the data.

Moreover, we created a barplot to highlight how apprehension levels deviate from
the average across months. The most striking observation is the sharp spike in 
March, followed by elevated levels in April and May, and then a decline into the
summer and fall months.

**Figure 4.** *Average Seasonal Pattern of Apprehensions by Calendar Month*
```{r seasonality 2,fig.align = 'center',fig.width=8, fig.height=5, echo = FALSE}
colours = ifelse(decomposition$figure > 0, "red", "gray")
barplot(decomposition$figure, names.arg = month.abb, col=colours)
```

This seasonal pattern likely reflects environmental realities faced by migrants. 
Crossing the U.S.–Mexico border involves walking through the desert, where 
extreme temperatures pose serious risks. The spring months offer a window 
of moderate weather, reducing the physical dangers associated with both winter
cold and summer heat.

## Correlations
In order to understand the broader determinants of migration flows, we analysed
whether U.S. border apprehensions correlate with economic and security variables
in Mexico, as well as with climatic variables in El Paso, Texas. The results
reveal nuanced and sometimes surprising insights.

### Economic Growth
To assess whether economic conditions in Mexico influence migration, I used the
*Indicador Global de la Actividad Económica (IGAE)* — a monthly index published
by *INEGI*, Mexico’s national statistics agency. The *IGAE* serves as a
high-frequency proxy for GDP, reflecting the country's overall economic 
performance by aggregating data from key sectors such as industry, services, and agriculture.

Despite the intuitive expectation that migration might increase during periods 
of economic distress, the correlation between monthly apprehensions and the 
*IGAE* index is zero (cor = 0.00026). As shown in this first scatter plot,
there is no visible relationship between the two variables. This suggests that
the national-level economic performance captured by *IGAE* may not have a direct
or immediate impact on **monthly** migration flows. Other forces, such as
insecurity or U.S. policy, may be more influential. It would be interesting to 
observe if, when transforming data to annual and using **actual** gdp data, this 
correlation grows.

**Figure 5.** *Apprehensions vs. Mexican Economic Activity (IGAE Index)*
```{r Economy, fig.align = 'center',fig.width=8, fig.height=5, echo = FALSE}
file_path <- file.path("Data", "Mexico's economic growth.csv")
MexEcon <- read.csv(file_path, header = TRUE)
MexEcon <- MexEcon[,c(1,6)] 
MexEcon <- na.omit(MexEcon)
colnames(MexEcon) <- c("Date", "Index")
MexEcon$Date <- as.Date(MexEcon$Date, format="%d/%m/%Y")
MexEcon$Date <- as.Date(MexEcon$Date,format="%Y/%m/%d" )

plot(apprehensionsTS, MexEcon$Index,
     xlab = "Monthly apprehensions in US' Southwest boarder",
     ylab = "Mexican exonomy (base: 01/01/2000)",
     pch = 19)
```

### Insecurity (Homicides in Mexico)
To explore the potential impact of violence on migration, we used monthly 
homicide statistics also published by *INEGI*. The data was retrieved through 
their multidimensional query system (*INEGI, OLAP Homicide Database*), which 
provides disaggregated information by month.

Contrary to what one might intuitively assume, there is a slightly negative 
correlation between monthly apprehensions and homicide counts (cor = -0.1163).
The second scatter plot shows that higher levels of violence do not necessarily 
lead to increased migration. One possible explanation is a *“frozen conflict”*
effect: in areas plagued by extreme cartel violence, people may be less able or
willing to leave due to fear, limited mobility, or blocked transit routes. 
In such contexts, the most vulnerable populations may become immobilized rather 
than mobilized — effectively trapped in violent environments.

This insight highlights that insecurity does not always result in higher
migration — especially when insecurity becomes so severe that it disrupts the 
ability to escape.

**Figure 6.** *Apprehensions vs. Homicide Counts in Mexico (INEGI)*
```{r insecurity, fig.align = 'center',fig.width=8, fig.height=5, echo = FALSE}
file_path <- file.path("Data", "Homicidios Mex.csv")
MexHom <- read.csv(file_path, header = TRUE)
MexHom <- MexHom[,c(1,4)] 
MexHom <- na.omit(MexHom)
colnames(MexHom) <- c("Date", "Homicides")
MexHom$Date <- as.Date(MexHom$Date, format="%d/%m/%Y")
MexHom$Date <- as.Date(MexHom$Date,format="%Y/%m/%d" )

plot(window(apprehensionsTS, start = c(2000, 1), end = c(2023, 12)),
     MexHom$Homicides,
     xlab = "Monthly apprehensions in US' Southwest boarder",
     ylab = "Number of monthly homocides in Mexico",
     pch = 19) 
```

### Climate (Average Monthly Temperature)
To examine whether climate conditions influence migration, we used monthly
average temperature data from El Paso, Texas — a key border crossing area. This 
dataset was sourced from the *National Centers for Environmental Information
(NCEI)* via *NOAA’s Climate at a Glance* portal.

The observed correlation between temperature and apprehensions was very weak and
slightly negative (cor = -0.0056). At first glance, this suggests temperature 
has little to no influence on migration. However, this result may be misleading
due to the nonlinear nature of the relationship.

It is possible that migration attempts decrease during both temperature 
extremes: when temperatures are too high, the desert crossing becomes 
life-threatening; when they are too low, especially during winter nights, the
risk of hypothermia also discourages movement. This would result in the highest
migration activity occurring during moderate spring temperatures — 
such as in March, when we observe seasonal peaks.

In such cases, the correlation appears flat not because there’s no relationship,
but because the relationship is U-shaped or threshold-based rather than linear.
A more accurate analysis could involve nonlinear models, temperature thresholds,
or categorizing the year into climate zones (e.g., *mild, hot, cold*) to 
assess their effects separately.

**Figure 7.** *Apprehensions vs. Average Monthly Temperature (El Paso)*
```{r climate, echo = FALSE}
file_path <- file.path("Data", "Weather.csv")
Climate <- read.csv(file_path, header = TRUE)
Climate$Date <- as.Date(paste0(Climate$Date, "01"), format = "%Y%m%d")
Climate$Value <- as.numeric(Climate$Value)*1000

FirstYear <- format(Climate$Date[1], "%Y")
FirstMonth <- format(Climate$Date[1], "%m")
ClimateTS <- ts(Climate$Value, start=c(FirstYear, FirstMonth), frequency=12)

tags$div(
  style = "text-align: center;",
  dygraph(cbind(apprehensionsTS, ClimateTS)) |>
  dySeries("apprehensionsTS", label = "Apprehensions") |>
  dySeries("ClimateTS", label = "Temperature (°F) * 1000") |>
  dyRangeSelector()
)
```

## Political Cycles and Presidential Effects
To better understand how U.S. political leadership may influence migration
patterns, we overlaid presidential transitions onto the apprehensions time 
series using vertical dashed lines. Each change in administration is marked with
its respective president's name and party color: red for Republicans (Bush,
Trump) and blue for Democrats (Obama, Biden).

**Figure 8.** *Apprehensions Over Time vs U.S. Presidential Administration Markers*
```{r presidents, echo=FALSE}
tags$div(
  style = "text-align: center;",
  dygraph(apprehensionsTS) |> dyRangeSelector() |> 
  dyEvent("2001-01-20", "George W. Bush", color="red") |>
  dyEvent("2009-01-20", "Barack Obama", color="blue") |>
  dyEvent("2017-01-20", "Donald Trump", color="red") |>
  dyEvent("2021-01-20", "Joe Biden", color="blue") |>
  dyEvent("2025-01-20", "Donald Trump", color="red")
)
```

### Visual Overview

The graph clearly shows distinct shifts in apprehension trends across 
presidencies:

- During **George W. Bush's** term (2001–2009), apprehensions declined 
significantly, continuing the downward trend from the early 2000s.

- Under **Barack Obama** (2009–2017), apprehensions remained relatively low and 
stable.

- Beginning with **Donald Trump’s** presidency in 2017, we observe an upward 
inflection in the trend, peaking around Spring 2019.

- During **Joe Biden’s** administration (2021–2025), apprehensions reached their 
highest recorded levels in the dataset.

### Statistical Comparison by Presidency
To test whether these visual differences were statistically meaningful, we 
calculated the mean number of apprehensions under each administration and 
constructed 95% confidence intervals:

- **Bush:** (79,302 – 84,773)

- **Obama:** (33,904 – 35,654)

- **Trump:** (30,519 – 37,082)

- **Biden:** (126,531 – 141,693)

As we can observe, Trump's and Obama's confidence intervals overlap, suggesting
no statistically significant difference in average apprehensions between their 
terms. This could be possibly due to the short duration of Trump’s presidency
(4 years) when compared to Obama's (8 years).

However, when comparing:

- **Before vs. After Trump:**

  - **Before:** (60,609 – 65,215)

  - **After:** (79,256 – 91,230)

we do observe a significant difference. And more importantly:

- **Before vs. After Biden:**

  - **Before:** (55,340 – 59,456)

  - **After:** (129,205 – 144,074)

There is a clear and statistically significant difference. The sharp rise in the
Biden era stands out as an **unprecedented shift**.

### Interpretation

This analysis suggests that while Trump’s rhetoric and border policies likely
had a long-term effect, **the largest increase in migration activity occurred 
under the Biden administration**. Despite President Biden’s campaign promises to
implement more humane immigration policies, his administration saw the highest 
number of border apprehensions in the dataset. This apparent contradiction can 
be explained by a combination of structural, policy, and contextual factors. 

Upon taking office, Biden rolled back several Trump-era enforcement measures, 
including ending the *Remain in Mexico* policy *(MPP)* and narrowing ICE
deportation priorities. These changes, while framed as humanitarian, may have 
been interpreted by migrants as a "green light", encouraging more attempts to 
cross the border. At the same time, an excess of migration demand that had been 
postponed during the COVID-19 pandemic surged in 2021, compounding the pressure 
at the border. Although the Biden administration retained Title 42
— a public health rule enabling immediate expulsions — the lack of legal 
consequences for those expelled led many migrants to **attempt multiple 
crossings** , artificially inflating apprehension figures. Ultimately, the spike
in apprehensions during Biden’s presidency reflects not only policy changes but
also a confluence of broader humanitarian crises and post-pandemic pressures.

It's important to interpret these results with nuance — **correlation does not 
imply causation** — but the data suggests that U.S. political leadership and 
enforcement strategies do impact migration flows, even if other structural 
forces also play a role.

## Conclusion
This analysis used monthly data from 2000 to early 2025 to examine the temporal 
dynamics of migrant apprehensions at the U.S. Southwest Border. Through a time
series framework, we identified strong seasonal patterns, a reversal in 
long-term trends, and significant structural breaks that coincide with political 
transitions and global shocks.

Our decomposition analysis confirmed a pronounced seasonality, with consistent
migration peaks in March, likely linked to favorable crossing conditions. The 
trend component revealed a long decline from 2000 to 2011, a stability through 
2017, and then an acceleration in apprehensions from there onward.

While national economic activity in Mexico *(IGAE)* showed no meaningful 
correlation with apprehensions, the data revealed a small negative correlation
with homicides, suggesting that extreme violence may immobilize rather than 
displace populations. Climate also showed a weak, non-linear relationship,
where migration likely declines in both temperature extremes, potentially
refering to a more complex dynamic that linear correlation fails to detect.

Political cycles introduced key structural changes. Although Trump's presidency 
marked a clear upward shift, the most statistically significant increase in 
apprehensions occurred under Biden, whose term aligned with post-pandemic surges
and changing enforcement strategies. Confidence interval analysis confirmed that
the mean number of apprehensions under Biden is significantly higher than under 
any prior administration in the dataset.

In summary, time series analysis reveals that **migration flows are shaped by the
interaction of seasonality, policy shifts, and external shocks**, rather than any
single driver. These findings highlight the utility of decomposition, 
correlation, and structural comparison techniques in disentangling short-term 
fluctuations from longer-term trends — and demonstrate the power of time series 
methods to explore politically and socially complex phenomena.

## Sources
- **U.S. Border Patrol Monthly Apprehensions**
U.S. Customs and Border Protection (CBP). Retrieved from:
https://www.cbp.gov/newsroom/stats/southwest-land-border-encounters

- **Mexico's Economic Activity (IGAE)**
Instituto Nacional de Estadística y Geografía (INEGI). Retrieved from:
https://www.inegi.org.mx/temas/igae/

- **Homicide Data in Mexico**
INEGI OLAP multidimensional query system. Retrieved from:
https://www.inegi.org.mx/sistemas/olap/proyectos/bd/continuas/mortalidad/defuncioneshom.asp?s=est#

- **Climate Data for El Paso, Texas**
National Centers for Environmental Information (NCEI), NOAA. Retrieved from:
https://www.ncei.noaa.gov/access/monitoring/climate-at-a-glance/city/time-series/USW00023044/tavg/1/0/2000-2025